---
title: "BioKG Demonstration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BioKG Demonstration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Download BioKG

The following Python script downloads the BioKG knowledge graph from the [Open Graph Benchmark (OGB)](https://ogb.stanford.edu/docs/linkprop/#ogbl-biokg).

```{python download-kg}
import numpy as np
import copy
import json
from ogb.linkproppred import LinkPropPredDataset

dataset = LinkPropPredDataset(name = "ogbl-biokg")

split_edge = dataset.get_edge_split()
train_edge, valid_edge, test_edge = split_edge["train"], split_edge["valid"], split_edge["test"]
graph = dataset[0] # graph: library-agnostic graph object
```

The edge index in the `graph` object is converted into JSON format to be read into R.

```{python edge-types}
print(graph.keys())
edge_index = graph["edge_index_dict"].copy()
```

To make the conversion, the dictionary keys must be renamed (i.e., cannot be tuples).

```{python rename-key}
old_keys = list(edge_index.keys())
for old_name in old_keys:
    new_name = "--".join(old_name)
    edge_index[new_name] = edge_index[old_name]
    del edge_index[old_name]
```

A special numpy encoder is defined, borrowed from [this StackOverflow post](https://stackoverflow.com/questions/57269741/typeerror-object-of-type-ndarray-is-not-json-serializable).

```{python numpy-encoder}
class NumpyEncoder(json.JSONEncoder):
    """ Special json encoder for numpy types """
    def default(self, obj):
        if isinstance(obj, (np.int_, np.intc, np.intp, np.int8,
                            np.int16, np.int32, np.int64, np.uint8,
                            np.uint16, np.uint32, np.uint64)):
            return int(obj)
        elif isinstance(obj, (np.float_, np.float16, np.float32,
                              np.float64)):
            return float(obj)
        elif isinstance(obj, (np.ndarray,)):
            return obj.tolist()
        return json.JSONEncoder.default(self, obj)
```

Finally, we convert the edge index to JSON and write to a file.

```{python convert-json}
edge_index_json = json.dumps(edge_index, cls = NumpyEncoder)

with open('inst//extdata//edge_index.json', 'a') as f:
    f.write(edge_index_json + '\n')
```

# Parse KG in R

Read the JSON dataset saved previously.

```{r read-kg}
# load libraries
library(data.table)
library(purrr)
library(magrittr)
library(rjson)

# load metapaths library
library(metapaths)

# read data
biokg = fromJSON(file = "inst/extdata/edge_index.json")
```

Read mappings downloaded from OGB.

```{r read-mappings}
mapping_dir = "inst/extdata/biokg_mappings"

read_mapping = function(mapping_path) {
  mapping = fread(file.path(mapping_dir, mapping_path))
  mapping[, Type := strsplit(mapping_path, "_")[[1]][1]]
  return(mapping)
}

mappings = list.files(mapping_dir) %>%
  .[grepl("entidx2name", .)] %>%
  map_dfr(read_mapping)

mappings = mappings %>% copy() %>%
  setnames(c("ent idx", "ent name"), c("Index", "Name")) %>%
  setcolorder("Type") %>%
  .[, Label := paste(Type, Index, sep = "_")]

```

Create a function to convert the edge list to a `data.table`. Note that the node IDs are specific to each type, so we must add a type-specific prefix.

```{r convert-biokg}
convert_biokg = function(sub_kg, sub_label) {
  
  # split label
  split_label = strsplit(sub_label, "--")[[1]]
  origin_label = paste(split_label[1], sub_kg[[1]], sep = "_")
  destination_label = paste(split_label[3], sub_kg[[2]], sep = "_")
  
  # create data table
  kg_dt = data.table(
    Origin = origin_label,
    Destination = destination_label,
    OriginType = split_label[1],
    DestinationType = split_label[3],
    EdgeType = split_label[2])
  
}
```

Now, map the conversion function over the `biokg` list.

```{r map-convert}
biokg_edge_list = imap_dfr(biokg, convert_biokg)
biokg_node_list = get_node_list(biokg_edge_list)
head(biokg_edge_list)
```

Check that the counts of each node type conform with `graph["num_nodes_dict"]` from the Python script.

| disease | drug  | function | protein | sideeffect |
|:-------:|:-----:|:--------:|:-------:|:----------:|
|  10687  | 10533 |  45085   |  17499  |    9969    |

```{r check-types}
biokg_node_list[, .N, by = NodeType]
```

Add node names from `mappings` table.

```{r add-name}
# add node names
biokg_node_list = merge(biokg_node_list, mappings[, .(Label, Name)], by.x = "Node", by.y = "Label", sort = F) %>%
  setnames("Name", "NodeName")
```

Sample the knowledge graph to generate a small, connected test set.

```{r sample-kg}
# load igraph library
library(igraph)

# generate graph
biokg_graph = igraph::graph.data.frame(biokg_edge_list, vertices = biokg_node_list, directed = T)

# sample random nodes
set.seed(42)
seed_nodes = sample(biokg_node_list$Node, 5000)

# generate induced subgraph
biokg_sample = igraph::induced.subgraph(graph = biokg_graph, vids = seed_nodes)

# get largest connected component
comp = igraph::components(biokg_sample) 
comp_idx = which.max(comp$csize)
lcc = comp$membership %>%
  {names(.)[. == comp_idx]}

# generate connected subgraph
biokg_sub = igraph::induced.subgraph(graph = biokg_graph, vids = lcc)

# generate subsamples node and edge lists
sub_edge_list = biokg_edge_list[Origin %in% lcc & Destination %in% lcc]
sub_node_list = biokg_node_list[Node %in% lcc]

# convenience function
lookup = function(node) {
  # return(sub_node_list[Node == node, NodeName])
  return(biokg_node_list[Node == node, NodeName])
}
```

Save node list and edge list to file.

```{r save-file}
save(sub_edge_list, file = "data/sub_edge_list.rda")
save(sub_node_list, file = "data/sub_node_list.rda")
```

# Test Node Similarity

Check possible edge types.

```{r check-edge}
message("- ", paste(unique(sub_node_list[, NodeType]), collapse = "\n- "))
```

Find hub genes.

```{r find-hubs}
# calculate degree
sub_degrees = degree(biokg_sub)

# find candidate nodes
candidate_nodes = sub_degrees %>%
  .[which(. > 20 & . < 30)] %>% names() %>% .[grepl("protein", .)] %>%
  { sub_node_list[Node %in% ., NodeName] }

cat(paste("HGNC:", candidate_nodes, sep = ""), sep = "\n")

```

Test similarity based on metapaths.

```{r test-similarity}
sim_test = get_similarity(
  "drug_762", "function_42893",
  mp = c("drug", "disease", "protein", "function"),
  node_list = biokg_node_list,
  edge_list = biokg_edge_list)

sim_res = sim_test$OriginPaths  %>%
  .[.[["function"]] == "function_42893", ]
```

Test negative similarity

```{r test-negative}
sim_neg_test = get_similarity(
  "drug_762", "function_29825",
  mp = c("drug", "disease", "protein", "function"),
  node_list = biokg_node_list,
  edge_list = biokg_edge_list)

sim_neg_res = sim_neg_test$OriginPaths %>%
  .[.[["function"]] == "function_29825", ]
```

Visualize results.

```{r visualize-results}
sim_nodes = unique(unlist(sim_res))

# generate connected subgraph
sim_graph = igraph::induced.subgraph(graph = biokg_graph, vids = sim_nodes)

# plot subgraph
plot(sim_graph, vertex.size = 15, edge.arrow.size = 0.25)
```

# Test Set Similarity

```{r test-aggregate}

alzheimer_drugs = c("drug_762", "drug_1075", "drug_895", "drug_1406")

alzheimer_pathways = c("function_36342", "function_26258", "function_21333", "function_37883", "function_42901", "function_17601", "function_36167")

test_agg = compare_sets(
  alzheimer_drugs,
  alzheimer_pathways,
  mp = c("drug", "disease", "protein", "function"),
  node_list = biokg_node_list,
  edge_list = biokg_edge_list)

```
