---
title: "Package Demonstration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package Demonstration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

First, load the `metapaths` package, as well as other requisite packages.

```{r setup}
# load packages
library(metapaths)
library(data.table)
library(magrittr)

# for comparison and visualization
library(igraph)
```

Print the example dataset.

```{r print-dataset}
mtcars_node_list = mtcars_edge_list[, .(Origin, OriginType)] %>%
  rbind(mtcars_edge_list[, .(Destination, DestinationType)], use.names = F) %>%
  unique() %>%
  setnames(c("Node", "NodeType"))

head(mtcars_node_list)
head(mtcars_edge_list)
```

Visualize the network as a graph.

```{plot-network}
mtcars_graph = graph.data.frame(mtcars_edge_list, vertices = mtcars_node_list, directed = T)
mtcars_col = factor(V(mtcars_graph)$NodeType)
plot(mtcars_graph, vertex.size = 15, vertex.label = NA, edge.arrow.size = .25, vertex.color = mtcars_col)
```

Get all node types.

```{r node-types}
mtcars_types = node_types(mtcars_edge_list, verbose = T)
```

Retrieve a list of unique nodes.

```{r all-nodes}
mtcars_nodes = all_nodes(mtcars_edge_list, verbose = T)
```

# Edge List

For a given node, find all neighbors of a specified type.

```{r get-neighbors}
get_neighbors("Cadillac Fleetwood", "Three", mtcars_edge_list) %>%
  cat(sep = "\n")
```

For a given node, find all neighbors of series of types.

```{r get-neighbors-type}
get_neighbors_type("Cadillac Fleetwood", c("Three", "Four"), mtcars_edge_list)
```

Construct the neighbor list.

```{r get-neighbor-list}
mtcars_neighbor_list = get_neighbor_list(mtcars_edge_list, verbose = T)
head(mtcars_neighbor_list)
```

# Neighbor List

Identify the neighbors using the constructed neighbor list rather than the edge list.

```{r search-neighbors}
search_neighbors("Cadillac Fleetwood", "Three", mtcars_neighbor_list)%>%
  cat(sep = "\n")
```

Compute degree stratified by node type. Here, degree is defined as the number of adjacent nodes (rather than edges).

```{r search-degrees}
search_degrees("Cadillac Fleetwood", "Three", mtcars_neighbor_list, "neighbor") %>%
  paste("Degree:", .) %>%
  cat(sep = "\n")
```

# Meta-Path Based Similarity

Compute similarity between two nodes using a sample meta-path: `Three-Three-Four-Five`.

```{r get-similarity}
mtcars_sim = get_similarity("Cadillac Fleetwood", "Ferrari Dino",
                            c("Three", "Three", "Four", "Five"),
                            c("pc", "npc", "dwpc"),
                            node_list = mtcars_node_list,
                            neighbor_list = mtcars_neighbor_list)
```

Let's inspect the output of the `get_similarity()` primitive. `get_similarity()` returns a named list. `OriginPaths` contains the paths from the origin node (i.e., `Cadillac Fleetwood`) to the destination node (i.e., `Ferrari Dino`) following the specified meta-path.

```{r origin-paths}
head(mtcars_sim$OriginPaths)
```

Similarly, `DestinationPaths` contains the paths from the destination node to the origin node following the reverse of the specified meta-path.

```{r destination-paths}
head(mtcars_sim$DestinationPaths)
```
